Thermal reactionome of a temperate ant species

### 2014-06-04 doc/ApTranscriptome_manuscript.Rmd
Updates to ms

When ants are exposed to different temperatures, lots of genes change expression. The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress.

Comparisons between species indicate that the species from the warmer environment tolerates thermal stress by higher baseline expression of protective genes

@@

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1). The density function for the temperature of maximum expression (Fig. 1) showed that the majority of responsive genes were in the *Low* category for both species, with *A. picea* having more *Low* genes than *A. carolinensis*. More genes had *Intermediate* expression in *A. carolinensis* than in *A. picea*, while the two species had a similar number of *High* genes (Fig. 1, Table 1).

@@

The reaction norms of thermally-responsive genes between species differed by mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis* than *A picea*, and mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about ...% (95% CI ... - ...%) and ...% (95% CI ... - ...%) were due to shifts in mean and slope, respectively.

We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for *High* genes 
*A. picea* (Wilcoxon V = `r w1$statistic`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r w1$conf.int[1:2]`), *Low* genes in *A. carolinensis* (Wilcoxon V = `r w2$statistic`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r w2$conf.int[1:2]`) and *Intermediate* genes in both (Wilcoxon V = `r w3$statistic`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r w3$conf.int[1:2]`), basal expression near optimum temperatures of 19.5C was always higher in *A. carolinensis*.

Overall greater thermal tolerance of one species would be indicated by finding that gene expression is stable across a greater portion of the thermal gradient. For *Bimodal* expressed genes, we found no difference (*t*~df=1406~ = 0.31, *P* = 0.76) between species in the variance of the reaction norms for thermally-responsive genes. In contrast, *Intermediate* genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In addition, the species differed in the temperatures at which thermally-responsive genes were activated. For genes with increased expression at *High* temperatures, we found that *A. carolinensis* genes were activated on average 1°C hotter than *A. picea* genes (Fig. 2). Similarly, for genes with increased expression at *Low* temperatures, *A. carolinensis* genes were activated on average 1°C hotter than *A. picea* genes (Fig. 2). Thus, while the activation temperatures differed, the size of the window between activation of *Low* and *High* genes remained the same between species, consistent with the results for *Bimodal* genes above.

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature"))
print(T_on_plot)
```

@@

overall pattern *A. carolinensis*

- more genes thermally-responsive
- tolerates thermal stress by shutting down biological processes

overall pattern *A. picea*

- more *Low* genes

- greater adaptability to cold than hot thermal stress
    - *A. carolinensis* limited future potential to adapt to heat stress
- different mechanisms of thermal tolernace
    - *A. carolinensis* tolerates thermal stress by shutting down (*Intermediate* genes)
    - *A. picea* tolerates by a combination of shutting down and activating 'protective' genes 
       - some indication that *A. picea* maintains express in cold conditions that *A. carolinensis* shuts down based on overlap of *Low* genes in *A. picea* with *Intermediate* genes in *A. carolinensis*
- metabolic compensation?

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [...], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis* found in colder and warmer environments, respectively. We found that nearly 10% of the genes in this species show significant thermal responsiveness.

### Molecular toolkit of thermal responsiveness

While the sheer number of thermally-responsive transcripts makes generalizations difficult, groupings such as Gene Ontology (GO) can be used to identify categories associated with thermal stress. For heat response, the semantic space of the enriched GO terms included *gene expression*, *regeneration* and *cellular component organization or biogenesis*. ...previous work... For cold response, there were 65 enriched GO terms that grouped by semantic space into *protein modification process*, *sensory perception of mechanical stimulus*, *response to mechanical stimulus*, *folic acid metabolism*, *membrane lipid metabolism* and *protein complex biogenesis*. ...previous work...

*Bimodal* genes are general stress response genes that are activated under both heat and cold stress. The semantic space of the enriched GO terms included *regulation of metabolism*, *biological regulation*, *macromolecule biosynthesis*, ...

*Intermediate* genes represent non-essential molecular functions that are shut-down as conditions become stressful. The semantic space for the enriched GO terms included *lipid metabolic process*, *glycoprotein biosynthetic process*, *membrane lipid metabolic process*, *cellular protein metabolic process*, *membrane lipid biosynthetic process*. ...indicates shutting down of non essential molecule biosynthesis...

**Sara - comparisons of our gene list to other studies**

This result poses a paradox. Why is the heat shock functional response highly conserved at the biological process level, while the exact molecular pathways appear to be highly labile? We suggest two possible explanations. First, regulatory networks controlling gene expression are evolutionarily dynamic. Trade-offs between function in different organisms may constrain which genes are involved in thermal response, thus resulting in different pathways to the same phenotype. Second, variation in molecular pathways may be due to inconsistent selection on thermal stress response. That is, as the environment changes and species shift their distributions, selection may relax or strengthen repeatedly within and across lineages. Different genes may be recruited within each selective episode. Differentiating between these, and alternate, explanations will require greater geographic and population genetic sampling.

### Transcriptome-wide comparisions

Of the 8,492 genes that were thermally-responsive, 80% differed in their responsiveness by colony. The majority of this difference in the reactionomes was due to changes in the slope ($\Delta S$) and curvature ($\Delta C$) which together accounted for nearly 90% of the differences between reaction norms for all transcripts. In a meta-analysis of phenotypic reaction norms, Murren et al. [-@murren2014] similarly found that differences in the shape of the reaction norms accounted for the majority of the differences. Thus, our finding suggests that this phenotypic pattern may be explained by greater genetic lability in the shape than the mean of gene expression. The relative role of plasticity versus adaptation in response to rapid environmental change is widely debated [@vellend...?]. Our data suggest that plasticity through changes in the responsiveness of gene expression are likely to occur more rapidly than changes in mean values.

An important note is that we would have completely missed this pattern if we had performed a differential expression study using only a few temperatures. Only by measuring gene expression across a thermal gradient and fitting a function to each transcript were we able to characterize the shape of the reaction norm of expression... To our knowledge, this is the first time that a reaction norm approach has been applied to the study of global gene expression.

Given the large differences in number (Table 1) and responsiveness of genes between the two species, a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures. Intriguingly, this result mirrors the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. A plausible explanation is that organisms have fewer molecular mechanisms to tolerate heat than cold stress. Future studies ... implications for climate change as the evolutionary potential of ecototherms to climate change may be limited as suggested by ... [kingsolver?]

@@

A key outcome of this study was developing the first genomic tools for the study of adaptive differentiation in *Aphaenogaster*. Our comparative results are limited to two populations, and an important future step will be expanding this work to examine differences in gene expression among a larger number of populations. Moreover, the potential for species to adapt to climate change will depend in part on their genetic variation for thermal tolerance. The quantitative genetic studies required to estimate this adaptive potential have been out of reach of ant biologists due to the long lifespan of species, and inability of ants to mate in the lab. However, future work could leverage genomic information to infer the relatedness among species in the absence of a known pedigree, enabling estimation of genetic variation within populations and their adaptive potential.

In summary, we have generated the first transcriptome of a temperate ant species, and have identified approximately 10% of genes that show thermal responsiveness. In this group, we find many known stress response genes, but also identify novel genetic pathways that respond to temperature. These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.

>>>>

- When ants are exposed to different temperatures, lots of genes change expression. 
- The number of responsive genes suggest that there is greater adaptability to cold than hot thermal stress. 
- Comparisons between species indicate that *A. carolinensis*, which has greater basal gene expression and physiological thermal tolerance, tolerates thermal stress by shutting down gene expression. *A. picea* tolerates thermal stress by activating protective genes

@@

We identified `r prettyNum(length(which(RxNpval$pval < 0.05)), big.mark = ",", scientific = FALSE)` genes with overall model fits of *P* < 0.05, retaining `r prettyNum(length(which(RxNpval$padj < 0.05)), big.mark = ",", scientific = FALSE)` after adjusting *P* values for false discovery rate. Of these genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` significantly change expression with temperature including `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` that have a species x temperature interaction. The number of transcripts that increased expression with temperature (*High*), decreased expression with temperature (*Low*), increased expression at both high and low (*Bimodal*) temperatures, decreased expression at high and low temperatures (*Intermediate*) or were not thermally-responsive in one species (*NotResp*) differed significantly between the species ($\chi^2$ = `r chi1$statistic`, P < `r max(chi1$p.value, 0.001)`; Table 1). The density function for the temperature of maximum expression (Fig. 1) showed that the majority of responsive genes were in the *Low* category for both species, with *A. picea* having more *Low* genes than *A. carolinensis*. More genes had *Intermediate* expression in *A. carolinensis* than in *A. picea*, while the two species had a similar number of *High* genes (Fig. 1, Table 1).

@@

The reaction norms of thermally-responsive genes between species differed by mean expression (*t*~`r t.mean$parameter`~ = `r t.mean$statistic`, *P* < `r max(t.mean$p.value, 0.001)`) which was about `r round(exp(t.mean$estimate))` TPM greater in *A. carolinensis* than *A picea*, and mean slope (*t*~`r t.slope$parameter`~ = `r t.slope$statistic`, *P* < `r max(t.slope$p.value, 0.001)`), which was about `r round(exp(t.mean$estimate))` $\Delta$TPM/degree greater in *A. picea*. Curvature (*t*~`r t.curvature$parameter`~ = `r t.curvature$statistic`, *P* < `r round(max(t.curvature$p.value, 0.001), 3)`) and wiggle (*t*~`r t.wiggle$parameter`~ = `r t.wiggle$statistic`, *P* < `r max(t.wiggle$p.value, 0.001)`) did not differ between species. Of the total differences ($\Delta T$) in reaction norms between species for each gene, about `r round(qlmean[2] * 100, 0)`% (95% CI `r paste(round(qlmean[1] * 100, 0), "--", round(qlmean[3] * 100, 0))`%) and `r round(qlslope[2] * 100, 0)`% (95% CI `r paste(round(qlslope[1] * 100, 0), "--", round(qlslope[3] * 100, 0))`%) were due to shifts in mean and slope, respectively.

We predicted that we would observe that genes up-regulated in response to thermal stress would have increased basal expression in the other species. Instead, we found that for genes expressed at *High* temperatures in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`), genes expressed at *Low* temperatures in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`) and *Intermediate* genes in both (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`), basal expression near optimum temperatures of 19.5°C was always higher in *A. carolinensis*.

Overall greater thermal tolerance of one species would be indicated by finding that gene expression is stable across a greater portion of the thermal gradient. For *Bimodal* expressed genes, we found no difference (*t*~df=1406~ = 0.31, *P* = 0.76) between species in the variance of the reaction norms for thermally-responsive genes. In contrast, *Intermediate* genes from *A. carolinensis* had a broader thermal range of stable gene expression than those from *A. picea* (*t*~df=1406~ = 9.25, *P* < 0.001, 95% CI difference in means 0.41 - 0.64). In addition, the species differed in the temperatures at which thermally-responsive genes were activated. For genes with increased expression at *High* temperatures, we found that *A. carolinensis* genes were activated on average at 33.8°C compared to 32.9°C for *A. picea* (Fig. 2). For genes with increased expression at *Low* temperatures, *A. carolinensis* genes were activated on average at 12.8°C compared to 13°C for a *A. picea* (Fig. 2).

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at *Low* and *High* temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}
T_on_plot <- ggplot(Ap.T_on, aes(x = T_on, y = ..density.., colour = colony, group = colony, fill = colony)) +
    geom_histogram(position = "dodge", binwidth = 3) +
    geom_density(alpha = 0.2, adjust = 3) + 
    scale_y_continuous(name="Density") +
    scale_x_continuous(name=expression("Temperature of gene activation"))
print(T_on_plot)
```

@@

- many genes thermally responsive
- thermal reaction norms differ between species
  - *A. carolinensis* limited future potential to adapt to heat stress
- more genes activated in response to cold shock --> greater adaptability to cold
- differences among species 
  - *A. carolinensis* tolerates thermal stress by shutting down biological processes
  - in *A. picea* the *Intermediate* genes from *A. carolinensis* are over-represented in *Low* and *Intermediate* categories
  - potential explanation is that *A. carolinensis* tolerates thermal stress by shutting-down metabolism and relies on other mechanisms (behavior, protein stability) whereas *A. picea* activates protective genes which requires metabolism...
- comparison of gene terms with other species (Sara?)

- regression approach allows us to detect patterns not evident from ANOVA design

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. While considerable research has focused on a few key genes (e.g. heat shock proteins) involved in thermal stress response [@krebs1997; @gaitainespitia2013], little is known about the functional diversity or transcriptome-wide patterns of thermal responsiveness. In this study, we characterized the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively. We found that nearly 10% of the genes in these species were thermally-responsive, with many more genes upregulated in response to cold than heat stress. The patterns of gene expression between species differed substantially, with more down-regulation of gene expression in *A. carolinensis* and more up-regulation of gene expression in *A. picea* in response to thermal stress (Fig. 2).

Given the differences in the number of thermally-responsive genes between the two species (Table 1), a striking pattern is that 3-4x more genes are upregulated at cold than high temperatures in both species, with greater variation between species at low temperatures (Fig. 1). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures. This explanation is consistent with the observation in *Aphaenogaster* [@warren2013] and other ectotherms [@addobediako2000; @hoffmann2013] that critical maximum temperatures are more constrained than critical minimum temperatures. Moreover, in an experiment with 94 species of *Drosophila*, Kellerman et al [-@kellerman2012] showed that limited genetic variation among species (e.g. phylogenetic inertia) rather than similar selection pressures underlie greater similarity of upper than lower thermal limits. Thus, our results indicate that, when coupled with phenotypic data, transcriptomic studies can provide insight on adapatability. For *Aphaenogaster*, our results suggest that adaptability to increased temperatures is limited, and these species risk replacement by more heat-tolerant competitors [@rodriguezcabal].

Depending on the frequency with which organisms experience stressful conditions, the reaction norms of protective genes may differ considerably. Organisms that frequently experience stress may have high basal levels of expression with little change, while organisms that less frequently experiences stress would have lower basal levels with large changes in response to environmental stress [@cite...]. For *Aphaenogaster*, we predicted that genes up-regulated to cold or hot temperatures in one species might be expressed at higher levels near optimal growth temperatures in the other species to confer greater basal stress tolerance. This prediction was not supported, as we found greater expression near optimal temperatures in *A. carolinensis* for *High* genes in *A. picea*, as predicted, but also for *Low* genes in *A. carolinensis* and *Intermediate* genes in both species. As *A. carolinensis* is from a warmer environment, a likely explanation for this is that this species simply has a higher metabolic rate and thus greater gene expression across it's thermal range. The observation that there are more *Intermediate* genes, which are shut-down under thermal stress, in *A. carolinensis* supports this explanation. ...cite some coral literature....

Of the `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes that were thermally-responsive, `r round(with(sigtable, sigtable[Coefficient=="Temperature:Colony","Number.significant"]) / with(sigtable, sum(sigtable[Coefficient %in% c("Temperature", "Temperature:Colony"),"Number.significant"])) * 100, 0)`% differed in their responsiveness by colony. The differences between species were due to changes in the mean ($\Delta M$) and slope ($\Delta S$) of the reaction norms for each thermally-responsive gene. This result contrasts with the findings of Murren et al. [-@murren2014], who found that most of the differences in phenotypic reaction norms were due to shape. As genes are not directly analogous to phenotypes, this contradiction is not surprising. Instead, it indicates that small shifts in the mean of many genes underlying a trait can result in non-linear phenotypic changes.

To date, transcriptomic studies have performed differential expression analysis between a few different conditions or environments (but see [@hodginsdavis2012]). By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. Thus, we were able to compare the temperature at which genes were activated near the upper and lower thermal limits between species (Fig. 2). For genes activated at cold temperatures, there was little difference among species with *A. picea* genes turning on at 13°C compared to 12.8°C for *A. carolinensis*. For high temperature genes, *A. carolinensis* genes were turned on at 33.8°C compared to 32.9°C for *A. picea*.

### Molecular toolkit of thermal tolerance

REANALYZING...

### Caveats/limitations

As with all *de novo* transcriptomes, our work has some important caveats. First, given that all sequenced ant genomes to date have about 18,000 genes [@smith...], our transcriptome is highly fragmented. Though we took steps to remove contaminants and redundant transcripts, some likely remain, in addition to partially-assembled transcripts. A genome assembly, in progress, would considerably resolve this problem. Second, the quality of the annotation for a non-model system such as *Aphaenogaster* is poor, as we have to rely upon resources from distantly related *Drosophila* and *Apis*. Finally, the mapping of changes in gene expression to organismal fitness is far from direct [@feder2005]. Large differences in patterns of gene expression may only have small effects on organismal fitness. Given these caveats, transcriptomic studies promise to offer many new insights on organisms' response to environmental change. Candidate gene studies, while informative, fail to capture the complexity of gene expression changes [@cite...]. In addition, for long-lived species that are intractable to study in the lab, transcriptomics can provide key insights on plasticity and the potential for species to tolerate and adapt to environmental change [@cite...].

@@

In this work, we have completed the first transcriptome for a temperate ant species, *Aphaenogaster picea* and *A. carolinensis*, to date. We found that a large proportion of the transcriptome is thermally-responsive. Transcriptome-wide changes in gene expression were greater at lower than upper thermal limits, indicating greater adapatabilty to cold than hot temperatures. ...differences among species...*A. carolinensis* greater organismal tolerance and shuts down metabolism under thermal stress...*A. picea* less tolerant and activates protective proteins requiring metabolism... These results provide key insights on the genetic basis of thermal tolerance, and provide a tremendous resource for the future study of ecological adaptation in ant species.
